The impact of gas bulk rotation on the Lyman-$\alpha$ line

### 2014-01-09 paper/ms.tex
update conclusions

\begin{abstract}
For the first time the effect of gas bulk rotation on the morphology 
of the Lyman $\alpha$ line is studied using radiative transfer calculations.
%We present results of radiative transfer calculations to measure
%the impact of gas bulk rotation on the morphology of the Lyman $\alpha$ line in galaxies. 
We model a galaxy as a sphere with a
homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with
maximum velocities in the range $0-300$ \kms and neutral
hydrogen optical depths in the range $\tau_{\rm
H}=10^{5}-10^{7}$. We also consider two kinds of spatial
distribution for the radiation sources with respect to the sphere:
central and homogeneous. We find that the line width and the intensity
at the line's center increases with rotational velocity. In the case of homogeneously distributed
sources, for large rotational velocities the line transforms from a
double peak to a single peak at the line center. Under the same
conditions the escape fraction increases $\sim 30\%$. For
radiation sources located off-center, the line morphology presents a
range of single, double and triple peaked lines. We show how these results are
useful to interpret recent spectroscopic results of distant $z\sim 2-3$ star
forming galaxies.
\end{abstract}

@@

In this paper, we have estimated the effects of gas bulk rotation on
the morphology of the \ly line in star forming
galaxies. The results are based on the study of an homogeneous sphere
with solid body rotation. We explored a range of models by varying
the rotational speed, hydrogen optical depth, dust optical depth and
initial distribution of \ly photons with respect to the gas
density. As a cross-validation, we obtain our results from two
independently developed Monte-Carlo radiative transfer codes.

Our main result is that rotation clearly impact the \ly line
morphology. Double peaked lines can make transitions into single
peaked lines as the rotational velocity increases. Additional
off-centered emission can result into triple peaked lines.

We have quantified the results in different aspects. In Section
\ref{sec:widthpeak} we show that the line width increases up to a
factor of $2-3$ in comparison to the static case. Simultaneously, the
flux in the line center also increases with rotational velocity. This
increase induces the transition from a double peak to a single peak
line in the models where the radiation sources are homogeneously
distributed.

In the case of homogeneously distributed sources, the changes in
line morphology are linked to the reduction in the average number of
scatterings. In the second part of Section \ref{sec:widthpeak} we show
that, as rotational velocity increases, a large fraction of photons
escape with less than $10$ scatterings. On average there are $40\%$
less scatterings. This makes highly improbable that a photon's
frequency can move away from the line's center.

In contrast, for centrally located sources, the number of scatterings
does not significantly decrease. However, due to the increasing
bulk velocity of the gas, each scattering is inefficient in shifting
photons away from the line's center.

These changes in the number of scatterings have a direct bearing on
the escape fraction in the presence of dust.  In Section
\ref{sec:escapefraction} we find that the escape fraction increases in
$20\%-30\%$ for the homogeneous source distribution. In the centrally
located emission, the rotational effects are negligible on the escape
fraction.

We also studied the effect of rotation in the position of the peaks in
the line. We find that in the regions with central emission the peaks
do not shift as rotation increases. However, in the homogeneous models
the double peak merges into a single peak.

@@

This paper illustrates for the first time the main effects of rotation
in the morphology of the \ly emission line, we estimate the
range of this effects for simplified models.

@@

JNGC thanks the support of Universidad de los Andes given as TA
Also ISAAC 2012 summer school which provide computer time where some of this 
models where ran.

>>>>

\begin{abstract}
%For the first the effect of gas bulk rotation on the morphology 
%of the Lyman $\alpha$ line is studied using radiative transfer calculations.
We present results of radiative transfer calculations to measure
the impact of gas bulk rotation on the morphology of the Lyman $\alpha$ line in galaxies. 
We model a galaxy as a sphere with a
homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with
maximum velocities in the range $0-300$ \kms and neutral
hydrogen optical depths in the range $\tau_{\rm
H}=10^{5}-10^{7}$. We also consider two kinds of spatial
distribution for the radiation sources with respect to the sphere:
central and homogeneous. We find that the line width and the intensity
at the line's center increases with rotational velocity. In the case of homogeneously distributed
sources, for large rotational velocities the line transforms from a
double peak to a single peak at the line center. Under the same
conditions the escape fraction increases $\sim 30\%$. For
radiation sources located off-center, the line morphology presents a
range of single, double and triple peaked lines. We show how these results are
useful to interpret recent spectroscopic results of distant $z\sim 2-3$ star
forming galaxies.
\end{abstract}

@@

This paper illustrates for the first time the main effects of rotation
in the morphology of the \ly emission line in star forming galaxies. 
The results are based on the study of an homogeneous sphere
with solid body rotation. We explored a range of models by varying
the rotational speed, hydrogen optical depth, dust optical depth and
initial distribution of \ly photons with respect to the gas
density. As a cross-validation, we obtain our results from two
independently developed Monte-Carlo radiative transfer codes.

Our main result is that rotation clearly impact the \ly line
morphology. Double peaked lines can make transitions into single
peaked lines as the rotational velocity increases. Additional
off-centered emission result into triple peaked lines for certain 
values of optical depth and velocities.

\begin{itemize}
\item For central sources we find that the number of scatterings
does not significantly decrease with rotation, this is due to the 
fact that the vast majority of scatterings events are resonant 
during which the mean free path is very short and the effect of gas
bulk rotation is not enough to affect \ly photons.
The escape fraction does not depend on the rotational velocity neither
the position of the peaks, but there is an increase 
in the flux on the central parts of the line for high rotational 
velocities.

\item For homogeneous sources we find the changes in the
line morphology are linked to the reduction in the average number of
scatterings. In the second part of \S \ref{sec:widthpeak} we show
that, as rotational velocity increases, a large fraction of photons
escape with less than $10$ scatterings and on average 
with $40\%$ less scatterings than in the static case. This makes highly improbable 
that a photon's frequency can move away from the line's center. 
In \S \ref{sec:widthpeak} For typical column densities of $N_{HI}=1^{19}-1^{20}$, 
we find that rotational velocities of 100-300 \kms broaden the line profile up 
to a factor of $2-3$ in comparison to the static case. Simultaneously, the
flux in the line center also increases with rotational velocity. This
increase induces the transition from a double peak to a single peak
line in the models where the radiation sources are homogeneously
distributed. Also the escape fraction increase by factors of $2-10$ with 
respect to the slab with centrally sources.

\item For non-symmetric sources we find that three peak profiles
could emerge for models in which the rotational velocity is the necessary 
to give rise to a central peak but also not that high to decrease the
amount of photons in the wings. When initial off-center sources are 
selected from the outer parts of the galaxy the wings of the line are stronger.
Also off-center emission do not produce any considerable asymmetries in the line.\\
\end{itemize}

%We have quantified the results in different aspects.

%These changes in the number of scatterings have a direct bearing on
%the escape fraction in the presence of dust.  In Section
%\ref{sec:escapefraction} we find that the escape fraction increases in
%$20\%-30\%$ for the homogeneous source distribution. In the centrally
%located emission, the rotational effects are negligible on the escape
%fraction.

%We also studied the effect of rotation in the position of the peaks in
%the line. We find that in the regions with central emission the peaks
%do not shift as rotation increases. However, in the homogeneous models
%the double peak merges into a single peak.

@@

@@

JNGC thanks the support of Universidad de los Andes given as teaching assistant, 
Also to the  ISAAC 2012 summer school in San Diego which provide computer time
 where some of this models where ran.
